---
title: "Co-flowering Synchrony and Bee Stability"
author: "M.D. Taylor & M. Chandar"
date: "2026-07-02"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
---
## Introduction
This R markdown contains all the analyses from the Co-flowering Synchrony & Bee Stability manuscript. This script contains 4 parts:

* Calculation of co-flowering synchrony using a circular statistics method
* Analysis 1: Changes in co-flowering synchrony over time
* Analysis 2: Climate predictors of co-flowering synchrony
* Analysis 3: Co-flowering synchrony and bee stability


This script requires the following data files (in order of use):

 * SEV phenology data
 * list of animal-pollinated SEV plants
 * SEV met49 and met50 climate data
 * Pacific Decadal Oscillation data 
 * SEV bee data

### Load packages 
```{r}
# Clear environment
rm(list = ls())

library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(ggplot2)
library(data.table)
library(Directional) #Note : Mac users may need to install "X-quartz" to use this package
library(sfsmisc)
library(scales)
library(reshape2)
library(data.table)
library(ggthemes)
library(cowplot)
library(ggpubr)
library(codyn)
library(nlme)
library(car)
library(visreg)
library(MuMIn)
library(emmeans)
library(piecewiseSEM)
library(nlme)
library(vegan)
library(plm)
library(lmerTest)
library(broom)

# Shared manuscript figure theme (full box, large bold fonts) + ecosystem palette
ms_theme <- ggplot2::theme_classic(base_size = 16) +
  ggplot2::theme(
    panel.border     = ggplot2::element_rect(fill = NA, colour = "black", linewidth = 1.1),
    axis.line        = ggplot2::element_blank(),
    axis.title       = ggplot2::element_text(face = "bold", size = 18),
    axis.text        = ggplot2::element_text(face = "bold", size = 13),
    strip.text       = ggplot2::element_text(face = "bold", size = 17),
    strip.background = ggplot2::element_blank(),
    legend.title     = ggplot2::element_text(face = "bold", size = 13),
    legend.text      = ggplot2::element_text(face = "bold", size = 12))
ms_site_colors <- c("B" = "#2c6c9c", "G" = "#8a8a8a", "C" = "#5f7d33")
```

### Load data
```{r load pheno data}
# SEV Phenology data
phen <- read.csv("sev137_plant_phenology.2023.csv", stringsAsFactors = FALSE) 
# List of animal-pollinated plants compiled by E. Martin
bee_poll_list<- read.csv("animal_poll_floral_async.csv") %>% rename(species = kartez)
```

## Calculating Synchrony (co-flowering overlap)
We used co-flowering overlap as a metric of floral synchrony. This method uses circular statistics to generate overlap values for all species pairs. This calculation is computationally intensive, so it is possible to skip to Analysis 1 and load in the resulting csv directly.

Wrangle Phenology Data and subset to animal-pollinated species 
```{r phen data wrangling}
# Ensure essential columns exist
required_cols <- c("date", "site", "web", "species", "reproduction")
stopifnot(all(required_cols %in% colnames(phen)))

# Remove NA months and unwanted sites
phen <- phen %>%
  filter(!is.na(date), site == "B"|site == "C"|site == "G") %>%
  mutate(
    year = as.integer(substr(date, 1, 4)),
    month = as.integer(substr(date, 6, 7)),
    day = as.integer(substr(date, 9, 10))
  )

#Keep only plant species that are animal-pollinated 
phen_bee <- phen %>% semi_join(bee_poll_list, by = "species")

# Keep only records with flowering reproduction codes (exclude "Z", "B", "FR")
# This gives us a data-set with abundance of plants that ARE IN FLOWER for each month
phen_flo <- phen_bee %>%
  filter(!reproduction %in% c("Z", "B", "FR"))


# Summarize flowering counts by species/site/year/month/web

# Detach plyr to avoid conflicts with dplyr
if ("package:plyr" %in% search()) detach("package:plyr", unload = TRUE)

# There are 5 webs at each site, which are surveyed each month within each year.
# This function summarizes the number of flowering individuals of a species in each web/site/month/year.
phen_flower <- phen_flo %>%
  group_by(year, month, site, web, species) %>%
  summarise(flospec_count = n(), .groups = "drop") %>%
  as.data.frame()

# Pivot wider 
wide_flower <- phen_flower %>%
  pivot_wider(
    names_from = species,
    values_from = flospec_count,
    values_fill = 0
  )

# Convert back to long format if needed
phen_flower_long <- wide_flower %>%
  pivot_longer(
    cols = -c(year, month, site, web),
    names_to = "species",
    values_to = "flospec_count"
  )

sort(unique(phen_flower_long$species))
```

Generate daily-resolution distributions
```{r von mises kernal density}
# Helper to compute circular kernel density estimate (von Mises)
densityFit <- function(x_rad, grid_rad, bandwidth) {
  # ensure input is numeric radians in [0, 2π]
  n <- length(x_rad)
  if (n == 0) return(rep(0, length(grid_rad)))
  
  # von Mises kernel density estimation manually
  res <- sapply(grid_rad, function(theta) {
    sum(exp(bandwidth * cos(theta - x_rad))) / (2 * pi * besselI(bandwidth, 0))
  })
  return(res)  # raw von Mises density; normalized to unit area in pheno.imputation
}

pheno.imputation <- function(rownum, ref_list, dat) {
  a <- ref_list[rownum, ]
  
  # Filter for the species/year/site/WEB combination
  sub <- dat %>%
    filter(species == a$species,
           year == a$year,
           site == a$site,
           web == a$web)
  
  # Skip if empty
  if (nrow(sub) == 0) return(NULL)
  
  # Skip if missing or all zero flowering counts
  if (all(is.na(sub$flospec_count)) || max(sub$flospec_count, na.rm = TRUE) == 0)
    return(NULL)
  
  # Map month to approximate DOY
  month_to_doy <- c(`3`=80, `4`=110, `5`=140, `6`=170, `7`=200,
                    `8`=230, `9`=260, `10`=290)
  sub <- sub %>%
    mutate(DOY = month_to_doy[as.character(month)]) %>%
    filter(!is.na(DOY))
  
  if (nrow(sub) == 0) return(NULL)
  
  # Count-shaped sample: repeat each month's DOY by its flowering count so the
  # kernel density reflects the WITHIN-species relative flowering intensity
  # across months. The density is normalized to unit area below, so between-
  # species magnitude (and the 10-individual censoring ceiling) drops out and
  # the overlap is the coefficient of overlapping (Ridout & Linkie 2009).
  lf <- rep(sub$DOY, times = sub$flospec_count)
  if (length(lf) == 0) return(NULL)
  
  # Convert to radians for circular KDE
  lf_rad <- (lf / 365) * 2 * pi
  day_rad <- (1:365 / 365) * 2 * pi
  
  # Compute von Mises kernel
  band <- vm.kde(lf_rad, thumb = "tay", rads = TRUE)$h
  if (is.nan(band) || is.null(band)) band <- 16.13
  
  circ_dist <- densityFit(lf_rad, day_rad, band)
  # Normalize to a unit-area probability density (Ridout & Linkie 2009) so that
  # overlap is the coefficient of overlapping, not an abundance-weighted metric
  circ_dist <- circ_dist / sum(circ_dist, na.rm = TRUE)
  
  data.frame(
    DOY = 1:365,
    abund = circ_dist,
    year = a$year,
    species = a$species,
    site = a$site,
    web = a$web,
    band = band
  )
}
```

Apply to all combinations
```{r apply to combos, eval=FALSE}
plant_list <- phen_flower_long %>%
  select(year, site, species, web) %>%
  distinct() %>%
  as.data.frame()

datalist <- list()
for (i in seq_len(nrow(plant_list))) {
  tryCatch({
    dat_i <- pheno.imputation(i, plant_list, phen_flower_long)
    if (!is.null(dat_i)) {
      datalist[[length(datalist) + 1]] <- dat_i
    }
    if (i %% 100 == 0) cat("Processed", i, "of", nrow(plant_list), "combinations\n")
  }, error = function(e) {
    message("Error at row ", i, ": ", e$message)
  })
}

# Combine
pheno_distributions <- data.table::rbindlist(datalist, fill = TRUE)
summary(pheno_distributions)
```

Generate species-species pairs per year/site/web
NOTE: 
 * There were 2 year-site-web groups in which only one species was present (and thus, co-flowering cannot be calculated) 
 * We used `phen_distrib_check %>% count(year, site, web, name = "n_species") %>% filter(n_species < 2)` to confirm the above
 * We removed these groups from the data set using `if (length(.x) < 2) return(NULL)`
 
```{r generate pairs, eval=FALSE}
phen_distrib_check <- pheno_distributions %>%
  select(year, site, web, species) %>%
  distinct()
summary(phen_distrib_check)

# Generate species-species pairs per year/site/web

plant_pairs <- phen_distrib_check %>%
  group_by(year, site, web) %>%
  summarise(species_list = list(species), .groups = "drop") %>%
  mutate(pairs = map(species_list, ~{
    if (length(.x) < 2) return(NULL) # ignores year/site/web grps with <2 species
    t(combn(.x, 2)) %>% as.data.frame() })) %>%
  select(-species_list) %>%
  unnest(pairs) %>%
  rename(species.x = V1, species.y = V2) %>%
  mutate(pair_id = pmap_chr(list(species.x, species.y),
        ~paste(sort(c(..1, ..2)), collapse = "_")),
         row = row_number()) %>%
  distinct(year, site, web, pair_id, .keep_all = TRUE)

summary(plant_pairs)
```

Calculate interspecific overlap
```{r calculate overlap, eval=FALSE}
inter_synchrony <- function(rownum, ref_pairs, dat) {
  a <- ref_pairs[rownum, ]

  sub1 <- dat %>% filter(species == a$species.x, year == a$year, site == a$site, web == a$web)
  sub2 <- dat %>% filter(species == a$species.y, year == a$year, site == a$site, web == a$web)

  d <- data.frame(
    DOY = sub1$DOY,
    species1 = sub1$abund,
    species2 = sub2$abund
  )

  d$min <- pmin(d$species1, d$species2)

  # Coefficient of overlapping (Delta): the shared area of the two unit-area
  # flowering densities (Weitzman 1970; Ridout & Linkie 2009). Because each
  # density integrates to 1, Delta ranges 0 (no overlap) to 1 (identical).
  delta <- integrate.xy(d$DOY, d$min)

  data.frame(
    year = a$year,
    species.x = a$species.x,
    species.y = a$species.y,
    site = a$site,
    web = a$web,
    overlap = delta
  )
}

# Loop through all pairs
datalist_overlap <- list()

library(furrr)
library(data.table)
library(sfsmisc)   # for integrate.xy
# 
# # Are all 5 transects sampled at each site every year? If yes, we are losing data somewhere 
# plant_pairs_test <- plant_pairs %>%
#   filter(year == 2010, site == "C", web == 2) %>%
#   slice_sample(n = 10)
# 
# test<- plant_pairs %>%
#   count(year, site, web)
# 
# 
# # Run the loop for just this subset
# datalist_overlap_test <- vector("list", length = nrow(plant_pairs_test))
# 
# for (i in 1:nrow(plant_pairs_test)) {
#   tryCatch({
#     dat <- inter_synchrony(i, plant_pairs_test, pheno_distributions)
#     if (!is.null(dat)) datalist_overlap_test[[i]] <- dat
#   }, error = function(e) {
#     message("Error in pair ", i, ": ", e$message)
#     NULL
#   })
# }
# 
# # Combine results
# plant_plant_overlap_test <- rbindlist(datalist_overlap_test, fill = TRUE)
# 
# # Inspect
# summary(plant_plant_overlap_test)
# 
# #####VISUALIZATION OF OVERLAP
# library(ggplot2)
# library(dplyr)
# 
# # Select the two species
# species1 <- plant_plant_overlap_test$species.x[3]
# species2 <- plant_plant_overlap_test$species.y[4]
# 
# # Filter phenology data
# pheno_plot_data <- pheno_distributions %>%
#   filter(year == 2010, 
#          site == plant_plant_overlap_test$site[1], 
#          species %in% c(species1, species2))
# 
# # Smooth curves for each species
# smooth_data <- pheno_plot_data %>%
#   group_by(species) %>%
#   arrange(DOY) # geom_smooth will handle smoothing
# 
# # Optional: compute overlap range (rough)
# overlap_data <- pheno_plot_data %>%
#   filter(species %in% c(species1, species2)) %>%
#   group_by(species) %>%
#   summarise(min_DOY = min(DOY[abund > 0]),
#             max_DOY = max(DOY[abund > 0])) %>%
#   summarise(overlap_start = max(min_DOY), overlap_end = min(max_DOY))
# 
# # Plot
# ggplot(pheno_plot_data, aes(x = DOY, y = abund, color = species)) +
#   geom_smooth(se = FALSE, span = 0.3, size = 1.2) +
#   # geom_rect(data = overlap_data, 
#             # aes(xmin = overlap_start, xmax = overlap_end, ymin = 0, ymax = 1),
#            # fill = "grey", alpha = 0.3, inherit.aes = FALSE) +
#   labs(title = paste("Phenology overlap:", species1, "and", species2),
#        x = "Day of Year", y = "Abundance") +
#   theme_minimal()

```

```{r run_overlap_parallel, eval=FALSE}
library(data.table)
library(future)
library(furrr)

n <- nrow(plant_pairs)
years <- plant_pairs$year
sites <- plant_pairs$site
webs <- plant_pairs$web
species_x <- plant_pairs$species.x
species_y <- plant_pairs$species.y

plan(multisession, workers = parallel::detectCores() - 1)

# Run overlap calculation in parallel, returning just a single numeric value per pair
overlaps <- future_map_dbl(1:n, function(i) {
  tryCatch({
    # Run inter_synchrony and extract only the 'overlap' column
    inter_synchrony(i, plant_pairs, pheno_distributions)$overlap
  }, error = function(e) NA_real_)
}, .progress = TRUE)

plan(sequential)

# Combine into a clean data.table
plant_plant_overlap <- data.table(
  year = plant_pairs$year,
  site = plant_pairs$site,
  web = plant_pairs$web,
  species.x = plant_pairs$species.x,
  species.y = plant_pairs$species.y,
  overlap = overlaps
)

str(plant_plant_overlap)
summary(plant_plant_overlap)

# Save output
write.csv(plant_plant_overlap, "plant_plant_overlapthru2024.csv", row.names = FALSE)
```
Now we have generated a csv containing the co-flowering overlap values of each species pair in a given year/site/web. 
We will use this data for all downstream analyses.

## Analysis 1: Co-flowering overlap over time
1.1 How has overall floral synchrony changed over time?
1.2 How has floral synchrony changed over time across three dryland ecosystem types?

### Prepare data
We will load in the co-flowering data from above and calculate average co-flowering overlap
```{r}
# Load co-flowering data
overlap <- read.csv("plant_plant_overlapthru2024.csv")
overlap <- overlap %>%
  filter(!site %in% c("J", "P", "JSAV"))

overlap$site <- as.factor(overlap$site)
overlap$species.x <- as.factor(overlap$species.x)
overlap$species.y <- as.factor(overlap$species.y)

# Calculate yearly average overlap for each transect
# Summarize yearly overlap per site per web_id over time
overlap <- overlap %>%
  dplyr:: group_by(site, year, web) %>%
  dplyr:: summarise(
    mean_overlap = mean(overlap, na.rm = TRUE),
    se_overlap = sd(overlap, na.rm = TRUE)/sqrt(n())
  ) %>%
  dplyr:: ungroup()
summary(overlap)

```

Load climate data and merge with co-flowering data
```{r}
# Load climate data (need to call dplyr)
PDO <- read.csv("pdo.timeseries.ersstv5.csv") %>% filter (year > 1998) %>% 
  dplyr:: group_by(year) %>% 
  dplyr:: summarize(mean(PDO)) %>% 
  dplyr:: rename(annual.PDO = `mean(PDO)`)
precip <- read.csv("climatebyyear_updated.csv") %>% select(-c(SPEI.comp.fall, SPEI.comp.spring))
precip.cv <- read.csv("precipitation_CV.csv")
precip <- merge(precip, precip.cv, by = c("site", "year"))
climate <-  merge(precip, PDO, by = "year", all.y = TRUE)
climate$annual.precip <- climate$season.precip.fall + climate$season.precip.spring 

# Assign met stations
overlap <- overlap %>%
  mutate(met_station = case_when(
    site %in% c("G", "C") ~ "met49",
    site == "B" ~ "met50"))

climate <- climate %>%
  dplyr::rename(met_station = site)

# Merge climate data with co-flowering data 
overlap <- merge(
  overlap,
  climate,
  by.x = c("year", "met_station"),
  by.y = c("year", "met_station"),
  all.x = TRUE) 

# add 2001 PDO back in for met50 2001 rows
overlap$annual.PDO[overlap$met_station == "met50" & overlap$year == 2001] <- -0.7419167

# Change web to web_id to make unique transect identifier (needed for AR models)
overlap <- overlap %>%
  mutate(web_id = interaction(site, web)) %>% 
  mutate(site = factor(site, levels = c("B", "G", "C"))) 

# Remove rows where se = NA, indicating only one overlap value
# Confirmed that this is the case to 2003 creosote outlier
overlap <- overlap %>% drop_na(se_overlap)

write.csv(overlap, "plant_overlap_climate_filtered.csv", row.names = FALSE)

#to read file back in, uncomment line below
#overlap<- read.csv("plant_overlap_climate_filtered.csv")
```

### 1.1 Change in synchrony over time (ALL SITES)
```{r}
# Q1: Is there a linear change in floral synchrony over time?
# two-stage: year is a landscape-level variable (all webs share the same
# year value), so lmer inflates DenDF to ~286 by treating annual rows as independent.
# alternatively we can extract one slope per web (Stage 1), then test slopes (Stage 2).
# This correctly makes DenDF 14 (global) and 12 (site contrast). Because transect (web) is the unit of replication

overlap_s <- overlap %>% mutate(year_scaled = as.numeric(scale(year)))

# Stage 1: one temporal slope per web (15 slopes)
web_slopes_year <- overlap_s %>%
  group_by(site, web_id) %>%
  do(tidy(lm(mean_overlap ~ year_scaled, data = .))) %>%
  filter(term == "year_scaled") %>%
  ungroup()

# Stage 2a: global trend — F_{1,14}
global_year_model <- lm(estimate ~ 1, data = web_slopes_year)
s <- summary(global_year_model)$coefficients
cat(sprintf("Global trend: F_{1,14} = %.3f, p = %.4f  (beta = %.4f, SE = %.4f)\n",
    s[1,"t value"]^2, s[1,"Pr(>|t|)"], s[1,"Estimate"], s[1,"Std. Error"]))
# F_{1,14} = 4.154, p = 0.061 — marginal, not significant

# Stage 2b: ecosystem contrast — F_{2,12}
site_year_model <- lm(estimate ~ site, data = web_slopes_year)
a_site <- anova(site_year_model)
cat(sprintf("Ecosystem contrast: F_{%d,%d} = %.3f, p = %.4f\n",
    a_site[1, "Df"], a_site[2, "Df"], a_site[1, "F value"], a_site[1, "Pr(>F)"]))

```

### 1.2 Change in synchrony over time by SITE
```{r}
#  per-site trends from the two-stage slopes above
# Each site has 5 webs F_{1,4} per site (5 webs - 1 = 4 df)

for (s in c("B","G","C")) {
  lab <- c(B="Plains Grassland", G="Desert Grassland", C="Desert Shrubland")
  sub <- web_slopes_year[web_slopes_year$site == s, ]
  m   <- lm(estimate ~ 1, data = sub)
  co  <- summary(m)$coefficients
  cat(sprintf("%s: F_{1,4} = %.3f, p = %.4f  (beta = %.4f, SE = %.4f)\n",
      lab[s], co[1,"t value"]^2, co[1,"Pr(>|t|)"], co[1,"Estimate"], co[1,"Std. Error"]))
}
# CORRECTED results:
#   Plains Grassland:  F_{1,4} = 0.882, p = 0.401
#   Desert Grassland:  F_{1,4} = 2.336, p = 0.201
#   Desert Shrubland:  F_{1,4} = 1.411, p = 0.301

```

### Figure 1 - Synchrony tracks PDO over time
```{r}
# PLOT synchrony over time with model fits 
site_colors <- c("B" = "deepskyblue4", "G" = "grey60", "C" = "darkolivegreen")
site_labels <- c("G" = "Desert Grassland", "B" = "Plains Grassland","C" = "Desert Shrubland")

# PDO ONLY PLOT
PDO.plot.raw <- ggplot(overlap, aes(x = year, y=annual.PDO)) +
  scale_linetype_manual(name = NULL, values = c("annual.PDO" = "dotted"), labels = c("annual.PDO"))+
  geom_line(data = overlap, aes(y = annual.PDO, linetype = "annual.PDO"),color = "black") +
  scale_y_continuous(
  name = "PDO", position = "right") +
  labs(x = "Year", y = "PDO") +
  theme_classic(base_size = 20) +
  facet_wrap(~ site,  ncol = 1, labeller = labeller(site = site_labels)) +
  guides(color = "none")
PDO.plot.raw
ggsave("FIGURE1_PDOraw.svg", PDO.plot.raw, height = 12, width = 14)
ggsave("FIGURE1_PDOraw.png", PDO.plot.raw, height = 12, width = 14, dpi = 300)


# OVERLAP ONLY PLOT
overlap.plot<- overlap %>% group_by(site, year, annual.PDO) %>% 
    summarise(
    mean_yr_overlap = mean(mean_overlap, na.rm = TRUE),
    se_overlap = sd(mean_overlap, na.rm = TRUE)/sqrt(n())
  ) %>% ungroup()

overlap.plot.raw <- ggplot(overlap.plot, aes(x = year, y = mean_yr_overlap)) +
  geom_point(aes(y = mean_yr_overlap, color = site), size = 2) +
  scale_linetype_manual(name = NULL, values = c("mean_yr_overlap" = "solid"), labels = c("mean_yr_overlap"))+
  geom_line(data = overlap.plot, aes(y=mean_yr_overlap, color = site, linetype = "mean_yr_overlap"))+
  scale_color_manual(values = site_colors, labels = site_labels) +
  geom_errorbar(aes(ymin = mean_yr_overlap - se_overlap, ymax = mean_yr_overlap + se_overlap, color = site), width = 0) +
  scale_y_continuous(
  name = "Co-flowering Overlap") +
  labs(x = "Year", y = "Mean Co-Flowering Synchrony") +
  theme_classic(base_size = 20) +
  facet_wrap(~ site,  ncol = 1, labeller = labeller(site = site_labels)) +
  guides(color = "none")
overlap.plot.raw
ggsave("FIGURE1_overlapraw.svg", overlap.plot.raw, height = 12, width = 14)
ggsave("FIGURE1_overlapraw.png", overlap.plot.raw, height = 12, width = 14, dpi = 300)

```
Combine raw plots in Adobe Illustrator to create dual-axes for figure 1.

## Analysis 2: Co-flowering overlap & Climate

### 2.1 Synchrony & Climate (ALL SITES)

MODEL SELECTION 1 - Temporal Autocorrelation
* Find best autocorrelation using linear model
* Next step uses best AR model to test non-linear models 
```{r}
oc <- overlap[!is.na(overlap$season.precip.spring) & !is.na(overlap$annual.PDO), ]
oc$year_fact <- as.factor(oc$year)

# FALL/MONSOON PRECIP
lm_overlap_fall_precip <- lme(mean_overlap ~ season.precip.fall * site, random = list(web_id = ~1, year_fact = ~1), method = "ML", data = oc, na.action = na.omit)
lm_overlap_fall_precip_AR1 <- lme(mean_overlap ~ season.precip.fall * site, random = ~1|web_id, method = "ML", correlation=corAR1(form=~year|web_id), data = oc, na.action = na.omit)
lm_overlap_fall_precip_AR2 <- lme(mean_overlap ~ season.precip.fall * site, random = ~1|web_id, method = "ML", correlation=corARMA(form=~year|web_id,p=2), data = oc, na.action = na.omit)
  
AIC.table1 <- model.sel(lm_overlap_fall_precip, lm_overlap_fall_precip_AR1, lm_overlap_fall_precip_AR2)
AIC.table1
# lm_overlap_fall_precip is the best model

# SPRING PRECIP
lm_overlap_spring_precip <- lme(mean_overlap ~ season.precip.spring * site, random = list(web_id = ~1, year_fact = ~1), method = "ML", data = oc, na.action = na.omit)
lm_overlap_spring_precip_AR1 <- lme(mean_overlap ~ season.precip.spring * site, random = ~1|web_id, method = "ML", correlation=corAR1(form=~year|web_id), data = oc, na.action = na.omit)
lm_overlap_spring_precip_AR2 <- lme(mean_overlap ~ season.precip.spring * site, random = ~1|web_id, method = "ML", correlation=corARMA(form=~year|web_id,p=2), data = oc, na.action = na.omit)

AIC.table2 <- model.sel(lm_overlap_spring_precip, lm_overlap_spring_precip_AR1, lm_overlap_spring_precip_AR2)
AIC.table2
# lm_overlap_spring_precip is the best model

# ANNUAL PRECIP
lm_overlap_annual_precip <- lme(mean_overlap ~ annual.precip * site, random = list(web_id = ~1, year_fact = ~1), method = "ML", data = oc, na.action = na.omit)
lm_overlap_annual_precip_AR1 <- lme(mean_overlap ~ annual.precip * site, random = ~1|web_id, method = "ML", correlation=corAR1(form=~year|web_id), data = oc, na.action = na.omit)
lm_overlap_annual_precip_AR2 <- lme(mean_overlap ~ annual.precip * site, random = ~1|web_id, method = "ML", correlation=corARMA(form=~year|web_id,p=2), data = oc, na.action = na.omit)

AIC.table3 <- model.sel(lm_overlap_annual_precip, lm_overlap_annual_precip_AR1, lm_overlap_annual_precip_AR2)
AIC.table3
# lm_overlap_annual_precip is the best model

# PDO
lm_overlap_PDO <- lme(mean_overlap ~ annual.PDO * site, random = list(web_id = ~1, year_fact = ~1), method = "ML", data = oc, na.action = na.omit)
lm_overlap_PDO_AR1 <- lme(mean_overlap ~ annual.PDO * site, random = ~1|web_id, method = "ML", correlation=corAR1(form=~year|web_id), data = oc, na.action = na.omit)
lm_overlap_PDO_AR2 <- lme(mean_overlap ~ annual.PDO * site, random = ~1|web_id, method = "ML", correlation=corARMA(form=~year|web_id,p=2), data = oc, na.action = na.omit)
AIC.table4 <- model.sel(lm_overlap_PDO, lm_overlap_PDO_AR1, lm_overlap_PDO_AR2)
AIC.table4
# lm_overlap_PDO_AR1 is the best model (delta = 4 w/ no AR)

# PRECIP CV
lm_overlap_precip_CV <- lme(mean_overlap ~ cv.daily.precip * site, random = list(web_id = ~1, year_fact = ~1), method = "ML", data = oc, na.action = na.omit)
lm_overlap_precip_CV_AR1 <- lme(mean_overlap ~ cv.daily.precip * site, random = ~1|web_id, method = "ML", correlation=corAR1(form=~year|web_id), data = oc, na.action = na.omit)
lm_overlap_precip_CV_AR2 <- lme(mean_overlap ~ cv.daily.precip * site, random = ~1|web_id, method = "ML", correlation=corARMA(form=~year|web_id,p=2), data = oc, na.action = na.omit)

AIC.table5 <- model.sel(lm_overlap_precip_CV, lm_overlap_precip_CV_AR1, lm_overlap_precip_CV_AR2)
AIC.table5
# lm_overlap_precip_CV is the best model

# Delta AIC does not differ that much between any except for PDO
```

MODEL SELECTION 2 - Non-linear fits using best models from above
Since almost none of the above best models had temporal autocorrelation, we opted to swap to lmer
```{r}
# Monsoon precip
lm_overlap_fall_precip <- lme(mean_overlap ~ season.precip.fall * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_fall_precip2 <- lme(mean_overlap ~ poly(season.precip.fall, 2) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_fall_precip3 <- lme(mean_overlap ~ poly(season.precip.fall, 3) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)

# Spring precip
lm_overlap_spring_precip <- lme(mean_overlap ~ season.precip.spring * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_spring_precip2 <- lme(mean_overlap ~ poly(season.precip.spring, 2) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_spring_precip3 <- lme(mean_overlap ~ poly(season.precip.spring, 3) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)

# Annual precip
lm_overlap_annual_precip <- lme(mean_overlap ~ annual.precip * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_annual_precip2 <- lme(mean_overlap ~ poly(annual.precip, 2) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_annual_precip3 <- lme(mean_overlap ~ poly(annual.precip, 3) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)

# PDO
lm_overlap_PDO <- lme(mean_overlap ~ annual.PDO * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_PDO2 <- lme(mean_overlap ~ poly(annual.PDO, 2) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_PDO3 <- lme(mean_overlap ~ poly(annual.PDO, 3) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)

# Precip CV
lm_overlap_precip_CV <- lme(mean_overlap ~ cv.daily.precip * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_precip_CV2 <- lme(mean_overlap ~ poly(cv.daily.precip, 2) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
lm_overlap_precip_CV3 <- lme(mean_overlap ~ poly(cv.daily.precip, 3) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)
# ================================
# Compare models with AICc
# ================================
AIC.table <- model.sel(lm_overlap_fall_precip, lm_overlap_fall_precip2,
                       lm_overlap_fall_precip3, lm_overlap_spring_precip,
                       lm_overlap_spring_precip2, lm_overlap_spring_precip3,
                       lm_overlap_annual_precip, lm_overlap_annual_precip2,
                       lm_overlap_annual_precip3, lm_overlap_PDO, lm_overlap_PDO2, lm_overlap_PDO3,
                       lm_overlap_precip_CV, lm_overlap_precip_CV2, lm_overlap_precip_CV3)
AIC.table
# Best model: SPRING precipitation (poly 2) is the best climate predictor of floral synchrony across sites!!
# Tied model: Spring precip (poly 3), highest ranked model (delta 1.4) but we went with simpler model
# 3rd best model: Monsoon (Fall) precip (poly 3) (delta 7.2)
# PDO (linear) is the 7th best model (delta 30)
```

Results from Spring precip and PDO models
crossed random effects lme() web_id absorbs between-transect baseline,
year_fact absorbs annual landscape-level climate clustering.
Site effect tested at DenDF = 12; climate effects at DenDF ≈ 269-272.
```{r}
# SPRING PRECIP — crossed random effects
best_model <- lme(mean_overlap ~ poly(season.precip.spring, 2) * site,
                  random = list(web_id = ~1, year_fact = ~1),
                  data = oc, method = "ML", na.action = na.omit)

aov_best <- anova(best_model, type = "marginal")
cat("Cool-season precipitation model (co-flowering synchrony ~ poly(precip, 2) * ecosystem):\n")
for (trm in rownames(aov_best)[-1]) {   # skip intercept
  cat(sprintf("  %-34s F_{%d,%d} = %.3f, p = %.4f\n", trm,
      aov_best[trm, "numDF"], aov_best[trm, "denDF"],
      aov_best[trm, "F-value"], aov_best[trm, "p-value"]))
}
# F_{2,269} = 8.86, p = 0.0002  (site: F_{2,12} = 2.08, p = 0.168)
# Interaction precip x site: F_{4,269} = 0.783, p = 0.537

emtrends(best_model, ~ site, var = "season.precip.spring", max.degree = 2)
# linear slopes (beta ± SE):
#   Plains Grassland (B): -0.0013 ± 0.0003
#   Desert Grassland (G): -0.0017 ± 0.0003
#   Desert Shrubland (C): -0.0019 ± 0.0003

qqnorm(resid(best_model))
hist(resid(best_model))

# PDO — crossed random effects
PDO_model <- lme(mean_overlap ~ annual.PDO * site,
                 random = list(web_id = ~1, year_fact = ~1),
                 data = oc, method = "ML", na.action = na.omit)

aov_pdo <- anova(PDO_model, type = "marginal")
cat("PDO model (co-flowering synchrony ~ PDO * ecosystem):\n")
for (trm in rownames(aov_pdo)[-1]) {   # skip intercept
  cat(sprintf("  %-34s F_{%d,%d} = %.3f, p = %.4f\n", trm,
      aov_pdo[trm, "numDF"], aov_pdo[trm, "denDF"],
      aov_pdo[trm, "F-value"], aov_pdo[trm, "p-value"]))
}
# F_{1,272} = 9.94, p = 0.0018  (site: F_{2,12} = 2.12, p = 0.163)
# Interaction PDO x site: F_{2,272} = 0.452, p = 0.637

emtrends(PDO_model, ~ site, var = "annual.PDO")
# PDO slopes (beta ± SE)
#   Plains Grassland (B): -0.0295 ± 0.0094
#   Desert Grassland (G): -0.0409 ± 0.0094
#   Desert Shrubland (C): -0.0398 ± 0.0094
```

## Figure 2: Synchrony, PDO, and Precipitation
```{r}
site_colors <- ms_site_colors
site_labels <- c("G" = "Desert Grassland", "B" = "Plains Grassland", "C" = "Desert Shrubland")

yearly_overlap <- overlap %>%
  dplyr::group_by(site, year) %>%
  dplyr::summarise(yearly_mean_overlap = mean(mean_overlap, na.rm = TRUE),
                   yearly_se_overlap   = sd(mean_overlap, na.rm = TRUE)/sqrt(n())) %>%
  dplyr::ungroup()
yearly_overlap <- merge(yearly_overlap,
  overlap %>% select(site, year, season.precip.fall, season.precip.spring, annual.PDO),
  by = c("site", "year")) %>% distinct(site, year, .keep_all = TRUE)

# across-site fit lines: cool-season precip (quadratic) and PDO (linear)
lm_spring_precip <- lme(mean_overlap ~ poly(season.precip.spring, 2),
                  random = list(web_id = ~1, year_fact = ~1), data = oc, method = "ML", na.action = na.omit)
fitA <- expand.grid(season.precip.spring = seq(min(overlap$season.precip.spring, na.rm = TRUE),
                    max(overlap$season.precip.spring, na.rm = TRUE), length.out = 1000))
fitA$pred_sync <- predict(lm_spring_precip, newdata = fitA, level = 0)

lm_PDO <- lme(mean_overlap ~ annual.PDO,
              random = list(web_id = ~1, year_fact = ~1), data = oc, method = "ML", na.action = na.omit)
fitB <- expand.grid(annual.PDO = seq(min(overlap$annual.PDO, na.rm = TRUE),
                    max(overlap$annual.PDO, na.rm = TRUE), length.out = 1000))
fitB$pred_sync <- predict(lm_PDO, newdata = fitB, level = 0)

panel_levels <- c("Cool season precipitation (mm)", "PDO")

long_clim <- yearly_overlap %>%
  tidyr::pivot_longer(c(season.precip.spring, annual.PDO),
                      names_to = "predictor", values_to = "xval") %>%
  mutate(panel = factor(dplyr::recode(predictor,
           season.precip.spring = "Cool season precipitation (mm)", annual.PDO = "PDO"),
           levels = panel_levels))

fitline <- rbind(
  data.frame(panel = "Cool season precipitation (mm)", xval = fitA$season.precip.spring, pred = fitA$pred_sync),
  data.frame(panel = "PDO", xval = fitB$annual.PDO, pred = fitB$pred_sync))
fitline$panel <- factor(fitline$panel, levels = panel_levels)

ann_let <- data.frame(panel = factor(panel_levels, levels = panel_levels), letter = c("A", "B"))

combo.plot <- ggplot(long_clim, aes(xval, yearly_mean_overlap, color = site)) +
  geom_errorbar(aes(ymin = yearly_mean_overlap - yearly_se_overlap,
                    ymax = yearly_mean_overlap + yearly_se_overlap),
                width = 0, linewidth = 0.7, alpha = 0.7) +
  geom_point(size = 2) +
  geom_line(data = fitline, aes(xval, pred), color = "black", linewidth = 1.4, inherit.aes = FALSE) +
  geom_text(data = ann_let, aes(x = -Inf, y = Inf, label = letter), inherit.aes = FALSE,
            hjust = -0.4, vjust = 1.2, size = 8, fontface = "bold") +
  scale_color_manual(values = site_colors, labels = site_labels, name = "") +
  facet_wrap(~panel, ncol = 1, scales = "free_x", strip.position = "bottom") +
  labs(x = NULL, y = "Co-flowering synchrony") +
  ms_theme +
  theme(
    legend.position      = c(0.99, 1.03),
    legend.justification = c(1, 1),
    legend.background    = element_blank(),
    legend.key           = element_blank(),
    strip.placement      = "outside",
    strip.background     = element_blank()
  )

combo.plot
ggsave("FIGURE2_ClimateSyn.svg", combo.plot, height = 10, width = 7.8)
ggsave("FIGURE2_ClimateSyn.png", combo.plot, height = 10, width = 7.8, dpi = 300)
```

### Figure S1 - Spring Precip over time
```{r}
# average cool-season precip of met49 and met50
climate_avg <- climate %>%
  group_by(year) %>%
  summarise(mean.precip.spring = mean(season.precip.spring, na.rm = TRUE),
            se.precip.spring   = sd(season.precip.spring, na.rm = TRUE)/sqrt(n())) %>%
  drop_na()

spring.precip.time.plot <- ggplot(climate_avg, aes(year, mean.precip.spring)) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = mean.precip.spring - se.precip.spring,
                    ymax = mean.precip.spring + se.precip.spring),
                width = 0, linewidth = 1, alpha = 0.75) +
  geom_point(size = 2.5) +
  labs(x = "Year", y = "Cool season precipitation (mm)") +
  ms_theme

spring.precip.time.plot
ggsave("FIGURES1_SpringPrecip.svg", spring.precip.time.plot, height = 10, width = 15)
ggsave("FIGURES1_SpringPrecip.png", spring.precip.time.plot, height = 10, width = 15, dpi = 300)

# Temporal trend in cool-season precipitation (reported in Figure S1 caption)
precip_time_lm <- lm(mean.precip.spring ~ year, data = climate_avg)
precip_time_sum <- summary(precip_time_lm)
precip_slope <- coef(precip_time_lm)[["year"]]
precip_F     <- precip_time_sum$fstatistic[["value"]]
precip_df1   <- precip_time_sum$fstatistic[["numdf"]]
precip_df2   <- precip_time_sum$fstatistic[["dendf"]]
precip_P     <- pf(precip_F, precip_df1, precip_df2, lower.tail = FALSE)
cat(sprintf("Cool-season precip vs year: beta = %.2f mm/yr, F_{%d,%d} = %.2f, P = %.3f, R2 = %.3f\n",
            precip_slope, precip_df1, precip_df2, precip_F, precip_P, precip_time_sum$r.squared))
```

## Analysis 3: Co-flowering overlap & Bee Stability 
### Prepare data
```{r}
# import full 2002 - 2019 bee abundance data 
bee.2019<-read.csv("SEVBeeData2002-2019_revised2023-07-19.csv",stringsAsFactors = T) 

# turn complete sampling month and complete sampling year into factors
bee.2019$complete_sampling_month <- as.factor(bee.2019$complete_sampling_month)
bee.2019$complete_sampling_year <- as.factor(bee.2019$complete_sampling_year)

# create unique identifier for each web
bee.2019$web_ID<-as.factor(bee.2019$transect)

# remove start/end date columns and collapse by month
bee.2019 <- bee.2019 %>% select (-c("start_date", "end_date", "direction", "color")) %>% 
  dplyr:: group_by(year, month, web_ID, ecosystem, complete_sampling_year) %>%  
  dplyr:: summarise(across(where(is.integer), sum))


# import Jade's tallied bee data (includes new data from 2020-2021 and some pre-2020 stragglers)
bee.tally<- read.csv("bee_tally.csv") %>% 
  separate(col = web_ID,
           into = c("web_ID", "direction"),
           sep = 2) %>%
  separate(col = web_ID,
           into = c("ecosystem", "web"),
           sep = 1, remove = FALSE) %>% 
  filter(!ecosystem %in% c("J", "P"))
         
# make wide 
bee.tally.wide <- bee.tally %>%  pivot_wider(
  id_cols = c(year, month, web_ID, ecosystem),
  names_from = code,
  values_from = abundance,
  values_fn = sum,
  values_fill = 0)

# merge tallied data with larger data set 
bee <- bind_rows(bee.2019, bee.tally.wide) %>%
  dplyr:: group_by(year, month, web_ID, ecosystem) %>%  
  dplyr:: summarise(across(where(is.integer), sum),
            .groups = "drop") %>% 
  dplyr::mutate(across(5:360, ~ replace_na(.x, 0)))
summary(bee)

# calculate total bee abundance
bee$abundance<-rowSums(bee[,5:360])
summary(bee$abundance)

# calculate richness, diversity, evenness
bee$richness<-specnumber(bee[,5:360])
bee$shannonH<-diversity(bee[,5:360])
bee$evenness<-bee$shannonH/log(bee$richness)

# import phenological synchrony
async<-read.csv("plant_overlap_climate_filtered.csv",stringsAsFactors = T) 

# create unique identifier for each web
async$web_ID<-as.factor(paste(async$site,async$web,sep=""))
bee$web_ID <- as.factor(bee$web_ID)
# check that they match
  summary(bee$web_ID) 
  summary(async$web_ID)

# remove 2016/2017 (partial sampling years)
bee<- bee %>% filter(!year %in% c("2016", "2017"))

# import M. Kazenel LH trait data
bee.spec <- read.csv("lhtraits_2023-08-29_forpub.csv")
#remove bees with unknown or parasitic diet breadth
bee.spec<- bee.spec %>% filter(diet_breadth != "unknown") %>%  filter(diet_breadth != "parasitic") %>% 
# some of the species have question marks  
filter(diet_breadth != "polylectic?") %>% filter(diet_breadth != "oligolectic?") 
sort(unique(bee.spec$diet_breadth))

# counts for number of gen vs specialist and total
bee.spec %>% 
  count(diet_breadth, sort = TRUE)
# out of 341 species: 86 specialists, 81 generalists 

```

### Calculating Bee Stability - POPULATION  
This represents population-level stability, calculating individual stability values for each species.

```{r}
# ============================================================================
# Bee stability over each species' ACTIVE WINDOW (MAIN) + whole season (SUPPL.)
# Rare species excluded (< 10 non-zero records). Window = one global, contiguous
# min-max non-zero month per bee species (across all years/ecosystems/transects);
# stability = 1 / CV of monthly abundance over those window months each year.
# ============================================================================

bee_long <- bee %>%
  select(-c(abundance, richness, shannonH, evenness)) %>%
  pivot_longer(cols = -c(year, ecosystem, web_ID, month),
               names_to = "species", values_to = "sp_abundance") %>%
  inner_join(bee.spec %>% select(code, diet_breadth), by = c("species" = "code"))

# Exclude truly rare species: keep those with >= 10 non-zero species x month records
keep_species <- bee_long %>%
  filter(sp_abundance > 0) %>%
  count(species, name = "n_rec") %>%
  filter(n_rec >= 10) %>%
  pull(species)
bee_long <- bee_long %>% filter(species %in% keep_species)

# One global, contiguous active window (min & max non-zero month) per species
sp_window <- bee_long %>%
  filter(sp_abundance > 0) %>%
  group_by(species) %>%
  summarise(min_mo = min(month), max_mo = max(month), .groups = "drop")

bee_wide <- bee_long %>%
  pivot_wider(id_cols = c(year, ecosystem, species, web_ID, diet_breadth),
              names_from = month, values_from = sp_abundance,
              values_fn = sum, values_fill = 0) %>%
  left_join(sp_window, by = "species")

month_cols <- as.character(sort(unique(bee_long$month)))
M  <- as.matrix(bee_wide[, month_cols]); storage.mode(M) <- "numeric"
mm <- as.integer(month_cols)

# stability = 1 / CV ; NA when < 2 months or zero total (CV undefined)
inv_cv <- function(v) {
  if (length(v) < 2 || sum(v) == 0) return(NA_real_)
  m <- mean(v); s <- sd(v)
  if (m == 0 || s == 0) return(NA_real_)
  1 / (s / m)
}

# MAIN: stability over each species' active window
bee_wide$bee.stability.win <- vapply(seq_len(nrow(bee_wide)), function(i) {
  inwin <- mm >= bee_wide$min_mo[i] & mm <= bee_wide$max_mo[i]
  inv_cv(M[i, inwin])
}, numeric(1))

# SUPPLEMENT: stability over the whole sampled season (all months)
bee_wide$bee.stability.full <- apply(M, 1, inv_cv)

# Average species-level stability to diet breadth per transect x year
agg_stability <- function(col) {
  bee_wide %>%
    filter(!is.na(.data[[col]])) %>%
    group_by(year, ecosystem, web_ID, diet_breadth) %>%
    summarise(bee.stability.sp = mean(.data[[col]]),
              bee.stability.se = sd(.data[[col]]) / sqrt(dplyr::n()),
              .groups = "drop") %>%
    mutate(diet_breadth = as.factor(diet_breadth))
}
bee_cast_sp   <- agg_stability("bee.stability.win")   # MAIN  (active window)
bee_cast_full <- agg_stability("bee.stability.full")  # SUPPLEMENT (whole season)
```

### 3.1 Bee Stability & Synchrony (ALL BEES)
```{r}
# Two-stage test (transect = unit of replication).
# Stage 1: slope of stability ~ synchrony per transect x diet breadth (with SE).
# Stage 2: test whether slopes differ from 0 (and by diet/ecosystem). Unweighted
# by default (each transect = one equally-informative replicate); set weighted=TRUE
# to down-weight imprecise transect slopes by 1/SE (Rudgers), as a robustness check.

two_stage_stability <- function(stab_df, label, weighted = FALSE) {
  m <- merge(async %>% select(year, web_ID, site, mean_overlap),
             stab_df, by = c("year", "web_ID"))
  m <- m %>% filter(web_ID != "C3")   # C3 excluded (insufficient data)
  m$ecosystem_type <- factor(
    dplyr::recode(m$ecosystem, B = "Plains grassland",
                  G = "Desert grassland", C = "Desert shrubland"),
    levels = c("Plains grassland", "Desert grassland", "Desert shrubland"))

  slopes <- m %>%
    group_by(web_ID, ecosystem_type, diet_breadth) %>%
    filter(dplyr::n() >= 3, dplyr::n_distinct(mean_overlap) >= 3) %>%
    do(tidy(lm(bee.stability.sp ~ mean_overlap, data = .))) %>%
    filter(term == "mean_overlap") %>%
    ungroup() %>%
    filter(!is.na(estimate), std.error > 0) %>%
    mutate(w = if (weighted) 1 / std.error else 1)   # 1/SE if weighted, else equal weights

  blab <- if (weighted) "1/SE-weighted mean β" else "mean β"
  cat("\n==== ", label, " ====\n", sep = "")
  for (d in unique(slopes$diet_breadth)) {
    s   <- slopes %>% filter(diet_breadth == d)
    fit <- lm(estimate ~ 1, weights = w, data = s)
    ci  <- confint(fit)
    tval <- summary(fit)$coefficients[1, "t value"]
    cat(sprintf("%-11s: n = %d, %s = %.3f [95%% CI %.3f, %.3f], F_{1,%d} = %.3f, P = %.4f\n",
        d, nrow(s), blab, coef(fit)[[1]], ci[1, 1], ci[1, 2],
        nrow(s) - 1, tval^2, summary(fit)$coefficients[1, 4]))
  }
  dc <- anova(lm(estimate ~ diet_breadth,   weights = w, data = slopes))
  ec <- anova(lm(estimate ~ ecosystem_type, weights = w, data = slopes))
  cat(sprintf("diet-breadth contrast: F = %.3f, P = %.4f\n", dc[1,"F value"], dc[1,"Pr(>F)"]))
  cat(sprintf("ecosystem contrast:    F = %.3f, P = %.4f\n", ec[1,"F value"], ec[1,"Pr(>F)"]))
  invisible(slopes)
}

# MAIN: active-window stability, UNWEIGHTED (each transect = one equal replicate)
two_stage_stability(bee_cast_sp,   "ACTIVE-WINDOW stability ~ synchrony (MAIN, unweighted)", weighted = FALSE)

# SUPPLEMENT 1: same analysis with 1/SE variance weighting -- shows the main
# result holds up when transect slopes are precision-weighted (Rudgers)
two_stage_stability(bee_cast_sp,   "ACTIVE-WINDOW stability ~ synchrony (SUPPL: 1/SE weighted)", weighted = TRUE)

# SUPPLEMENT 2: whole-season stability (all Mar-Oct months; plant-community view)
two_stage_stability(bee_cast_full, "WHOLE-SEASON stability ~ synchrony (SUPPL: whole season)", weighted = FALSE)
```

## Figure 3: Generalist vs. Specialist Bee Stability vs. overlap
```{r}
# Figure 3: bee stability vs co-flowering synchrony -- unified manuscript style.
make_stability_fig <- function(stab_df, outfile) {
  d <- merge(async %>% select(year, web_ID, site, mean_overlap),
             stab_df, by = c("year", "web_ID")) %>% filter(web_ID != "C3")
  d$ecosystem_type <- factor(
    dplyr::recode(d$ecosystem, B = "Plains grassland", G = "Desert grassland", C = "Desert shrubland"),
    levels = c("Plains grassland", "Desert grassland", "Desert shrubland"))
  d$diet_lab <- factor(
    dplyr::recode(as.character(d$diet_breadth), polylectic = "Generalists", oligolectic = "Specialists"),
    levels = c("Generalists", "Specialists"))

  # Stage-1 slopes: one per transect (web_ID) within diet x ecosystem
  slopes <- d %>% group_by(diet_lab, ecosystem_type, web_ID) %>%
    filter(dplyr::n() >= 3, dplyr::n_distinct(mean_overlap) >= 3) %>%
    do(tidy(lm(bee.stability.sp ~ mean_overlap, data = .))) %>%
    filter(term == "mean_overlap") %>% ungroup()

  # Global (diet-level) mean slope = mean of transect slopes -> bold black line
  stat <- slopes %>% group_by(diet_lab) %>%
    summarise(b = mean(estimate), n = dplyr::n(), se = sd(estimate)/sqrt(dplyr::n()),
              tcrit = qt(0.975, dplyr::n() - 1), p = t.test(estimate)$p.value, .groups = "drop") %>%
    mutate(lo = b - tcrit*se, hi = b + tcrit*se)

  centro <- d %>% group_by(diet_lab) %>%
    summarise(x0 = mean(mean_overlap), y0 = mean(bee.stability.sp),
              xmin = min(mean_overlap), xmax = max(mean_overlap), .groups = "drop") %>%
    left_join(stat, by = "diet_lab")
  bold <- centro %>% group_by(diet_lab) %>%
    do({ cc <- .; xs <- seq(cc$xmin, cc$xmax, length.out = 100)
         data.frame(diet_lab = cc$diet_lab, mean_overlap = xs,
                    y  = cc$y0 + cc$b*(xs - cc$x0),
                    lo = cc$y0 + cc$b*(xs - cc$x0) - cc$tcrit*cc$se*abs(xs - cc$x0),
                    hi = cc$y0 + cc$b*(xs - cc$x0) + cc$tcrit*cc$se*abs(xs - cc$x0)) }) %>% ungroup()

  # Per-ecosystem mean slope = mean of that ecosystem's transect slopes (matches
  # the per-ecosystem beta values reported in the Figure 3 caption), drawn through
  # each ecosystem's centroid over its own synchrony range.
  eco_stat <- slopes %>% group_by(diet_lab, ecosystem_type) %>%
    summarise(b_eco = mean(estimate), .groups = "drop")
  eco_centro <- d %>% group_by(diet_lab, ecosystem_type) %>%
    summarise(x0 = mean(mean_overlap), y0 = mean(bee.stability.sp),
              xmin = min(mean_overlap), xmax = max(mean_overlap), .groups = "drop") %>%
    left_join(eco_stat, by = c("diet_lab", "ecosystem_type"))
  eco_lines <- eco_centro %>% group_by(diet_lab, ecosystem_type) %>%
    do({ cc <- .; xs <- seq(cc$xmin, cc$xmax, length.out = 100)
         data.frame(diet_lab = cc$diet_lab, ecosystem_type = cc$ecosystem_type,
                    mean_overlap = xs, y = cc$y0 + cc$b_eco*(xs - cc$x0)) }) %>% ungroup()

  ann <- stat %>% mutate(
    letter = ifelse(diet_lab == "Generalists", "A", "B"))

  cols <- c("Plains grassland" = "#2c6c9c", "Desert grassland" = "#8a8a8a", "Desert shrubland" = "#5f7d33")
  p <- ggplot(d, aes(mean_overlap, bee.stability.sp)) +
    geom_point(aes(color = ecosystem_type), size = 1.8, alpha = 0.55) +
    geom_line(data = eco_lines, aes(x = mean_overlap, y = y, color = ecosystem_type, group = ecosystem_type),
              inherit.aes = FALSE, linewidth = 1) +
    geom_ribbon(data = bold, aes(x = mean_overlap, ymin = lo, ymax = hi), inherit.aes = FALSE, fill = "black", alpha = 0.12) +
    geom_line(data = bold, aes(x = mean_overlap, y = y), inherit.aes = FALSE, linewidth = 1.4) +
    geom_text(data = ann, aes(x = -Inf, y = Inf, label = letter), inherit.aes = FALSE,
              hjust = -0.4, vjust = 1.2, size = 8, fontface = "bold") +
    scale_color_manual(values = cols, name = NULL) +
    facet_wrap(~diet_lab, ncol = 1) +
    labs(x = "Co-flowering synchrony", y = "Bee stability (1/CV)") +
    ms_theme + theme(legend.position = c(0.99, 0.99), legend.justification = c(1, 1))
  p
}
make_stability_fig(bee_cast_sp,   "FIGURE3_stability_MAIN_activewindow.svg")
make_stability_fig(bee_cast_full, "FIGURE3_stability_SUPPL_wholeseason.svg")
```




